Simple repression draft

### 2020-06-02 doc/section_01_introduction.tex
MRM pass through intro and discussion to add more citations

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures. This challenge stems first and foremost from our ignorance about
what those architectures even are with more than 60\% of the genes even in an
ostensibly well understood organism such as  {\it E. coli} with no regulatory
insights at all~\cite{Rydenfelt2014-2,Belliveau2018,Ghatak2019,
Santos_Zavaleta2019}. But even once we have established the identity of key
transcription factors and their binding sites of a given promoter architecture,
there remains the predictive challenge of understanding its input-output
properties, an objective that can be met by a myriad of approaches using the
tools of statistical physics~\cite{Ackers1982, Shea1985,
Buchler2003,Vilar2003a,Vilar2003b, Bintu2005a,Bintu2005c, Gertz2009,Sherman2012,
Saiz2013, Ko1991,Peccoud1995,Record1996, Kepler2001,
Sanchez2008,Shahrezaei2008,Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists.
Our aim is to understand the underlying principles of such coarse-graining
in order to increase the predictive power of these models. In doing so we hope
future models of regulatory response will be able to serve the powerful
predictive role needed to take synthetic biology from a brilliant exercise in
enlightened empiricism to a rational design framework as  in any other branch of
engineering.
% Our aim is to understand the underlying principles of such
% coarse-graining such that future models of regulatory response can serve the
% powerful predictive role needed to take synthetic biology from what is often a
% brilliant exercise in enlightened empiricism and recast it as the same kind of
% rational design process we are used to in other branches of engineering. 
More precisely, we want phenomenology in the sense of coarse-graining away
atomistic detail, but still retaining biophysical meaning. For example, we are
not satisfied with the strictly phenomenological approach offered by the
commonly used Hill functions. Studies like~\cite{Razo-Mejia2018} have
demonstrated that Hill functions are clearly insufficient since each new
situation requires a completely new set of parameters. Such work requires a
quantitative theory of how biophysical changes at the molecular level propagate
to input-output functions at the genetic circuit level. We want concepts, not
mere facts. In particular a key question is: at this level of coarse graining,
what microscopic details do we need to explicitly model, and how do we figure
that out? For example, do we need to worry about all or even any of the steps
that individual RNAPs go through each time they make a transcript? Turning the
question around, can we see any imprint of those processes in the available
data? If the answer is no, then those processes are irrelevant for our purposes.
Forward modeling and inverse (statistical inferential) modeling are necessary to 
tackles such questions.

Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple
repression that is implicit in the repressor-operator model. An operator, the
binding site on the DNA for a repressor protein, may be found occupied by a
repressor, in which case transcription is blocked from occurring. Alternatively,
that binding site may be found unoccupied, in which case RNA polymerase (RNAP)
may bind and transcription can proceed. The key assumption we make in this
simplest incarnation of the repressor-operator model is that binding of
repressor and RNAP in the promoter region of interest is exclusive, meaning that
one or the other may bind, but never may both be simultaneously bound. It is
often imagined that when the repressor is bound to its operator, RNAP is
sterically blocked from binding to its promoter sequence. Current evidence
suggests this is sometimes, but not always the case, and it remains an
interesting open question precisely how a repressor bound far upstream is able
to repress transcription. \marginpar{\it Cite that fig, from ??Nathan or Bill or
Rob??, showing locations of repressor binding sites far upstream of promoter}
Suggestions include ``action-at-a-distance'' mediated by kinks in the DNA,
formed when the repressor is bound, that prevent RNAP binding. Nevertheless, our
modeling in this work is sufficiently coarse-grained that we simply assume
exclusive binding and leave explicit accounting of these details out of the
problem.

>>>>

Gene expression presides over much of the most important dynamism of living
organisms.   The level of expression of batteries of different genes is altered
as a result of spatiotemporal cues that integrate chemical, mechanical and other
types of signals.  The original repressor-operator model conceived by Jacob and
Monod in the context of bacterial metabolism has now been transformed into the
much broader subject of gene regulatory networks in living organisms of all
kinds~\cite{Jacob1961, Britten1969, Ben-TabouDe-Leon2007}.  One of the remaining
outstanding challenges to have emerged in the genomic era is our continued
inability to predict the regulatory consequences of different regulatory
architectures, i.e. the arrangement and affinity of binding sites for
transcription factors and RNA polymerases on the DNA. This challenge stems
first and foremost from our ignorance about what those architectures even are
with more than 60\% of the genes even in an ostensibly well understood organism
such as  {\it E. coli} with no regulatory insights at all~\cite{Rydenfelt2014-2,
Belliveau2018, Ghatak2019, Santos_Zavaleta2019}. But even once we have
established the identity of key transcription factors and their binding sites of
a given promoter architecture, there remains the predictive challenge of
understanding its input-output properties, an objective that can be met by a
myriad of approaches using the tools of statistical physics~\cite{Ackers1982,
Shea1985, Buchler2003, Vilar2003a, Vilar2003b, Bintu2005a, Bintu2005c,
Gertz2009,Sherman2012, Saiz2013, Ko1991, Peccoud1995, Record1996, Kepler2001,
Sanchez2008, Shahrezaei2008, Sanchez2011, Michel2010}.  One root to such
predictive understanding is to focus on the simplest regulatory architecture and
to push the theory-experiment dialogue as far and as hard as it can be
pushed~\cite{Phillips2019}.  If we demonstrate that we can pass that test by
successfully predicting both the means and variance in gene expression, then
that provides a more solid foundation upon which to launch into more complex
problems.

To that end, in this paper we examine a wide variety of distinct models for the
simple repression regulatory architecture. This genetic architecture consists of
a DNA promoter regulated by a transcriptional repressor that binds to a single
binding site \cite{Garcia2011a}. All of the proposed models models coarse-grain
away some of the important microscopic features of this architecture that have
been elucidated by generations of geneticists, molecular biologists and
biochemists. Our aim is to understand the underlying principles of such
coarse-graining in order to increase the predictive power of these models. In
doing so we hope future models of regulatory response will be able to serve the
powerful predictive role needed to take synthetic biology from a brilliant
exercise in enlightened empiricism to a rational design framework as  in any
other branch of engineering. More precisely, we want phenomenology in the sense
of coarse-graining away atomistic detail, but still retaining biophysical
meaning. For example, we are not satisfied with the strictly phenomenological
approach offered by the commonly used Hill functions~\cite{Frank2013}.
Studies like~\cite{Razo-Mejia2018} have demonstrated that Hill functions are
clearly insufficient since each new situation requires a completely new set of
parameters. Such work requires a quantitative theory of how biophysical changes
at the molecular level propagate to input-output functions at the genetic
circuit level. We want concepts, not mere facts. In particular a key question
is: at this level of coarse graining, what microscopic details do we need to
explicitly model, and how do we figure that out? For example, do we need to
worry about all or even any of the steps that individual RNAPs go through each
time they make a transcript? Turning the question around, can we see any imprint
of those processes in the available data? If the answer is no, then those
processes are irrelevant for our purposes. Forward modeling and inverse
(statistical inferential) modeling are necessary to tackles such questions.

Figure~\ref{fig1:means_cartoons}(A) shows the qualitative picture of simple
repression that is implicit in the repressor-operator model. An operator, the
binding site on the DNA for a repressor protein, may be found occupied by a
repressor, in which case transcription is blocked from occurring. Alternatively,
that binding site may be found unoccupied, in which case RNA polymerase (RNAP)
may bind and transcription can proceed. The key assumption we make in this
simplest incarnation of the repressor-operator model is that binding of
repressor and RNAP in the promoter region of interest is exclusive, meaning that
one or the other may bind, but never may both be simultaneously bound. It is
often imagined that when the repressor is bound to its operator, RNAP is
sterically blocked from binding to its promoter sequence. Current evidence
suggests this is sometimes, but not always the case, and it remains an
interesting open question precisely how a repressor bound far upstream is able
to repress transcription~\cite{Rydenfelt2014-2}. Suggestions include
``action-at-a-distance'' mediated by kinks in the DNA, formed when the repressor
is bound, that prevent RNAP binding. Nevertheless, our modeling in this work is
sufficiently coarse-grained that we simply assume exclusive binding and leave
explicit accounting of these details out of the problem.

### 2020-06-02 doc/section_05_discussion.tex
MRM pass through intro and discussion to add more citations

The study of gene expression is one of the dominant themes of modern biology,
made all the more urgent by the dizzying pace at which genomes are being
sequenced. But there is a troubling Achilles heel buried in all of that genomic
data and that is our inability to find and interpret regulatory sequence. In
many cases, this is not possible even qualitatively, let alone the possibility
of quantitative dissection of the regulatory parts of genomes in a predictive
fashion. Other recent work has tackled the challenge of finding and annotating
the regulatory part of genomes \cite{Belliveau2018, Ireland2020} \mrm{add other
non-self referential}. Once we have determined the architecture of the
regulatory part of the genome, we are then faced with the next class of
questions which are sharpened by formulating them in mathematical terms, namely,
what are the input-output properties of these regulatory circuits and what knobs
control them?

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod. Regulation in that architecture is
the result of a competition between a repressor which inhibits transcription and
RNAP polymerase which undertakes it. Through the labors of generations of
geneticists, molecular biologists and biochemists, an overwhelming amount of
information and insight has been garnered into this simple regulatory motif,
licensing it as what one might call the ``hydrogen atom'' of regulatory biology.
It is from that perspective that the present paper explores the extent to which
some of the different models that have been articulated to describe that motif
allow us to understand both the average level of gene expression found in a
population of cells, the intrinsic cell-to-cell variability, and the full gene
expression distribution found in such population as would be reported in a
single molecule Fluorescence in situ Hybridization experiment, for example.

Our key insights can be summarized as follows. First, as shown in
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (variance/mean) that are less than one, at odds with any and all
experimental data that we are aware of. This theoretical result allows us to
directly discard a subset of the models (models 1-3 in
Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
experimental observations. The two remaining models (models 4 and 5 in
Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic
realistm to be able to reproduce the data. A previous exploration of one of
these models demonstrated the sloppy nature of the model in which data on
single-cell mRNA counts alone cannot constrain the value of all parameters
simultaneously \cite{Razo-Mejia2020}. Here we demonstrate that the proposed
one-state bursty promoter model (model 5 in Figure~\ref{fig2:constit_cartoons})
emerges as  a limit of the commonly used two-state promoter model
\cite{Peccoud1995}. We put  the idea to the test that this level of
coarse-graining is rich enough to reproduce previous experimental observations.
In particular we perform Bayesian inference to determine the two parameters
describing the full steady-state mRNA distribution, finding that the model is
able to provide a quantitative description of  a plethora of promoter sequences
with different mean levels of expression and noise.

With the results of the constitutive promoter in hand, we then fix the
parameters associated with this class of promoters and use them as input for
evaluating the noise in gene expression for the simple repression motif itself.
This analysis allows us to provide a single overarching analysis of both the
constitutive and simple repression architectures using one simple model and
corresponding set of self-consistent parameters, demonstrating not only a
predictive framework, but also reconciling the equilibrium and non-equilibrium
views of the same simple repression constructs.

It is interesting to speculate what microscopic details are being coarse-grained
by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons} model 5.
Chromosomal locus is one possible influence we have not addressed in this work,
as all the FISH data from~\cite{Jones2014} that we considered was from a
construct integrated at the \textit{galK} locus. The results of~\cite{Chong2014}
indicate that transcription-induced supercoiling contributes substantially in
driving transcriptional bursting, and furthermore, their Figure~7 suggests that
the parameters describing the rate, duration, and size of bursts vary
substantially for transcription from different genomic loci. Although the
authors of~\cite{Englaender2017} do not address noise, they note enormous
differences in mean expression levels when an identical construct is integrated
at different genomic loci. The authors of~\cite{Engl2020} attribute noise and
burstiness in their FISH data to the influence of different sigma factors, which
is a reasonable conclusion from their data. Could the difference also be due to
the different chromosomal locations of the two operons?

@@

\mmnote{Do all my approximations work simply because galK happens to have a low
duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
interesting test: put in the dozen or so Hernan constructs at some other locus,
preferably one of the ones Chong identified as having a large duty cycle (large
$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
fold-changes. You don't even need to do FISH, although that would be even
better. If their model is right, $\rho$ should be approximately set by the duty
cycle of supercoiling, NOT by RNAP kinetics.
\textit{This is a NICE prediction for future work.} \textit{cspE}
is claimed nonessential, prediction is you drop HG reporter constructs in there
and the inferred binding energies of all lacI operators shift the same amount,
by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
confounding effects from moving to a new locus. With FISH measurements, $k^+$
and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
measurements of $\beta/\alpha$ are correct. This would require an obnoxious
amount of cloning, but once you did that, the measurements would be same old
stuff that RP lab does.}

@@

>>>>

The study of gene expression is one of the dominant themes of modern biology,
made all the more urgent by the dizzying pace at which genomes are being
sequenced. But there is a troubling Achilles heel buried in all of that genomic
data and that is our inability to find and interpret regulatory sequence. In
many cases, this is not possible even qualitatively, let alone the possibility
of quantitative dissection of the regulatory parts of genomes in a predictive
fashion. Other recent work has tackled the challenge of finding and annotating
the regulatory part of genomes \cite{Belliveau2018, Ireland2020}. Once we have
determined the architecture of the regulatory part of the genome, we are then
faced with the next class of questions which are sharpened by formulating them
in mathematical terms, namely, what are the input-output properties of these
regulatory circuits and what knobs control them?

The present work has tackled that question in the context of the first
regulatory architecture hypothesized in the molecular biology era, namely, the
repressor-operator model of Jacob and Monod~\cite{Jacob1961}. Regulation in that
architecture is the result of a competition between a repressor which inhibits
transcription and RNAP polymerase which undertakes it. Through the labors of
generations of geneticists, molecular biologists and biochemists, an
overwhelming amount of information and insight has been garnered into this
simple regulatory motif, licensing it as what one might call the ``hydrogen
atom'' of regulatory biology. It is from that perspective that the present paper
explores the extent to which some of the different models that have been
articulated to describe that motif allow us to understand both the average level
of gene expression found in a population of cells, the intrinsic cell-to-cell
variability, and the full gene expression distribution found in such population
as would be reported in a single molecule Fluorescence in situ Hybridization
experiment, for example.

Our key insights can be summarized as follows. First, as shown in
Figure~\ref{fig1:means_cartoons}, the mean expression in the simple repression
architecture is captured by a master curve in which the action of repressor and
the details of the polymerase interaction with the promoter appear separately
and additively in an effective free energy. Interestingly, as has been shown
elsewhere in the context of the Monod-Wyman-Changeux model, these kinds of
coarse-graining results are an exact mathematical result and do not constitute
hopeful approximations or biological naivete \cite{Razo-Mejia2018, Chure2019}.
To further dissect the relative merits of the different models, we must appeal
to higher moments of the gene expression probability distribution. To that end,
our second set of insights focus on gene expression noise, where it is seen that
a treatment of the constitutive promoter already reveals that some models have
Fano factors (variance/mean) that are less than one, at odds with any and all
experimental data that we are aware of\cite{So2011, Jones2014}. This theoretical
result allows us to directly discard a subset of the models (models 1-3 in
Figure~\ref{fig2:constit_cartoons}(A)) since they cannot be reconciled with
experimental observations. The two remaining models (models 4 and 5 in
Figure~\ref{fig2:constit_cartoons}) appear to contain enough microscopic realism
to be able to reproduce the data. A previous exploration of model 4 demonstrated
the sloppy nature of the model in which data on single-cell mRNA counts alone
cannot constrain the value of all parameters simultaneously
\cite{Razo-Mejia2020}. Here we demonstrate that the proposed one-state bursty
promoter model (model 5 in Figure~\ref{fig2:constit_cartoons}) emerges as  a
limit of the commonly used two-state promoter model \cite{Peccoud1995,
Shahrezaei2008, So2011, Sanchez2013, Jones2014}. We put the idea to the test
that this level of coarse-graining is rich enough to reproduce previous
experimental observations. In particular we perform Bayesian inference to
determine the two parameters describing the full steady-state mRNA distribution,
finding that the model is able to provide a quantitative description of a
plethora of promoter sequences with different mean levels of expression and
noise.

With the results of the constitutive promoter in hand, we then fix the
parameters associated with this class of promoters and use them as input for
evaluating the noise in gene expression for the simple repression motif itself.
This allows us to provide a single overarching analysis of both the constitutive
and simple repression architectures using one simple model and corresponding set
of self-consistent parameters, demonstrating not only a predictive framework,
but also reconciling the equilibrium and non-equilibrium views of the same
simple repression constructs. More specifically, we obtained values for the
transcription factor association and dissociation rates by performing Bayesian
inference on the ful mRNA distribution for data obtained from simple-repression
promoters with varying number of transcription factors per cell and affinity of
such transcription factors for the binding site. The free energy value obtained
from these kinetic rates -- computed as the log ratio of the rates -- agrees
with previous inferences performed only from mean gene expression measurements,
that assumed an equilibrium rather than a kinetic
framework~\cite{Garcia2011a,Razo-Mejia2018}.

It is interesting to speculate what microscopic details are being coarse-grained
by our burst rate and burst size in Figure~\ref{fig2:constit_cartoons} model 5.
Chromosomal locus is one possible influence we have not addressed in this work,
as all the single-molecule mRNA data from~\cite{Jones2014} that we considered
was from a construct integrated at the \textit{galK} locus. The results
of~\cite{Chong2014} indicate that transcription-induced supercoiling contributes
substantially in driving transcriptional bursting, and furthermore, their
Figure~7 suggests that the parameters describing the rate, duration, and size of
bursts vary substantially for transcription from different genomic loci.
Although the authors of~\cite{Englaender2017} do not address noise, they note
enormous differences in mean expression levels when an identical construct is
integrated at different genomic loci. The authors of~\cite{Engl2020} attribute
noise and burstiness in their single-molecule mRNA data to the influence of
different sigma factors, which is a reasonable conclusion from their data. Could
the difference also be due to the different chromosomal locations of the two
operons?

@@

\mrm{We need to decide ASAP if this is going to be included in the discussion or not. My vote is no, but Jupiter must decide.}
\mmnote{Do all my approximations work simply because galK happens to have a low
duty cycle of supercoiling, $\beta/\alpha$ in Chong2014 picture? A potentially
interesting test: put in the dozen or so Hernan constructs at some other locus,
preferably one of the ones Chong identified as having a large duty cycle (large
$\beta/\alpha$), i.e., the opposite limit of galK. Then just measure
fold-changes. You don't even need to do FISH, although that would be even
better. If their model is right, $\rho$ should be approximately set by the duty
cycle of supercoiling, NOT by RNAP kinetics.
\textit{This is a NICE prediction for future work.} \textit{cspE}
is claimed nonessential, prediction is you drop HG reporter constructs in there
and the inferred binding energies of all lacI operators shift the same amount,
by about $\ln(3)\sim 1kT$. Not huge, but detectable, although might be
confounding effects from moving to a new locus. With FISH measurements, $k^+$
and $k^-$ in Fig 2, model 4 should now be separately inferrable if Chong2014
measurements of $\beta/\alpha$ are correct. This would require an obnoxious
amount of cloning, but once you did that, the measurements would be same old
stuff that RP lab does.}

@@

\mrm{Rather than proposing more experiments, I feel that we are missing a closing paragraph that puts all these findings in context. Where do we go from here as a community? Things such as the next building block, i.e. simple activation, and then from there scale up the complexity until we can get a full gene regulatory network kind of thing. Rather than going small into tiny details specific to this, let's step back and look at the big picture.}
